---
title: 'Introduction to R, Week 1: Functions and objects'
output: github_document
---

<!--week1.md is generated from week1.Rmd. Please edit that file -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Objectives

Welcome to Introduction to R from fredhutch.io!
This course introduces you R by working through activities common in data science:
importing, manipulating, and visualizing data. 

By the end of this week's session, you should be able to:

- work within the RStudio interface to run and save R code in a project
- understand basic R syntax to use functions and assign objects
- create and manipulate vectors and understand how R deals with missing data 

## A brief orientation to RStudio 

motivation for R and RStudio

- large user community
- RStudio makes working with R much easier
- create new project in new directory (recommended to place on Desktop or in Documents)
- create new R script, append with .R
- top left window: source (script), where commands are saved
- bottom left window: console, where commands are executed
- top right window: environment, showing what "things" are "remembered" by R
- bottom right window: file browser, plots, packages, help
- many buttons and keyboard shortcuts available for working in RStudio
- workflow: testing commands in source, using keyboard shortcut to send to console

other keyboard shortcuts available in Help menu

we have access to RStudio server through http://rstudio.fhcrc.org/

## Using functions



```{r}
# basic math
4 + 5 
```

```{r}
# same thing works without spaces!
4+5
```

Brief overview to style is [here](http://adv-r.had.co.nz/Style.html),
more information is available in the [tidyverse style guide](https://style.tidyverse.org).

```{r}
# using a function
round(3.14)
# finding help for a command
args(round)
```

`?round`

```{r}
# using a function with more arguments
round(3.14, digits = 1)
```


```{r}
# can switch order of arguments (but need to name value!)
round(digits = 1, x = 3.14)
```

**Challenge:** what does the function hist do? What are its main arguments? How did you determine this?

To wrap R source files: Tools -> Global Options -> Code, check box for "Soft-wrap R source files"

## Assigning objects

```{r}
# assigning value to an object
weight_kg <- 55 # <- is an assignment operator
# recall object
weight_kg
```

```{r}
# perform an operation on an object
2.2 * weight_kg
```

```{r}
# assign a new value to object
weight_kg <- 57.5
```

```{r}
# create new object from operation
weight_lb <- 2.2 * weight_kg
```

```{r}
weight_kg <- 100
```

The order in which operations are executed matters!

```{r}
# remove object
remove(weight_kg) # can also abbreviate to rm
```

**Challenge:** what is the value of each item at each step?
```{r}
mass <- 47.5            # mass?
width  <- 122             # width?
mass <- mass * 2.0      # mass?
width  <- width - 20        # width?
mass_index <- mass/width  # mass_index?
```

## Vectors

```{r}
# assign vector
ages <- c(50, 55, 60, 65) # c for combine/concatenate
# recall vector
ages
```


```{r}
# how many things are in object?
length(ages)
# what type of object?
class(ages)
# get overview of object
str(ages)
```


```{r}
# performing functions with vectors
mean(ages)
range(ages)
```


```{r}
# vector of characters
organs <- c("lung", "prostate", "breast")

```

**Challenge:**

- How many things are in organs?
- what type of data is organs?
- get overview of organs

Data types in R

- "character" for strings, uses quotes
-  "numeric" (or "double")
-  "integer" for integer numbers (e.g., 2L, the L indicates to R that it’s an integer)
- "logical" for TRUE and FALSE (the boolean data type)
- "complex" to represent complex numbers with real and imaginary parts (e.g., 1 + 4i)
- "raw" for bitstreams

**Challenge:** what happens when each of the objects are created?

```{r}
num_char <- c(1, 2, 3, "a")
num_logical <- c(1, 2, 3, TRUE)
char_logical <- c("a", "b", "c", TRUE)
tricky <- c(1, 2, 3, "4")
```

## Working with vectors

```{r}
# add a value at the end of vector
ages <- c(ages, 90) # same combine function as creating vector
```

```{r}
# add value at the beginning
ages <- c(30, ages)
```

```{r}
# subsetting vectors (index starts at 1)
organs[2] # extracting single value
```


```{r}
organs[c(1, 3)] # extracting multiple values
```

```{r}
organs[-2] # excluding values
```

```{r}
# conditional subsetting
ages > 60 # identifies whether each part of vector meets condition
```

```{r}
ages[ages > 60] # extracts values which meet condition
```

```{r}
ages[ages == 60]
```

Can also include <= and >=

```{r}
# combining conditions: OR
ages[ages < 50 | ages > 60]
```

# can also combine condition with AND &, but be careful about vernacular
# OR means value must meet one or more conditions
# AND means value must meet both conditions

```{r}
ages[ages < 50 & ages > 60] # not possible!

```

**Challenge:** why does the following code return TRUE: 
```{r}
"four" > "five"
```

## Missing data

```{r}
# create new vector with missing data indicated by NA
heights <- c(2, 4, 4, NA, 6)
```

```{r}
# perform function on vector
mean(heights) # what is wrong with this answer?
max(heights)
```

```{r}
# add argument to remove NA
mean(heights, na.rm = TRUE)
max(heights, na.rm = TRUE)
```

Additional ways to remove missing data

```{r}
# identify elements which are missing data
is.na(heights)
```

```{r}
# reverse the TRUE/FALSE
!is.na(heights)
```

```{r}
# extract elements which are not missing values
heights[!is.na(heights)]
```

```{r}
# remove incomplete cases
na.omit(heights) # what is different about this result?
```

**Challenge:**
```{r}
# create vector
more_heights <- c(63, 69, 60, 65, NA, 68, 61, 70, 61, 59, 64, 69, 63, 63, NA, 72, 65, 64, 70, 63, 65)
```

- remove NAs from `more_heights`
- calculate median
- identify how many elements from more_heights are greater than 67 inches
- visualize data as a histogram (hint: function `hist`)

## Wrapping up

- review objectives
- direct towards practice questions (linked in HackMD)
- preview next week's objectives
- closing RStudio projects:
-   don't save .RData
-   can change in Global Options

If you would like more practice, 
the [`exercises`](exercises/) directory contains additional tasks for practice.
Answers to those and in-class questions are available in the [`solutions`](solutions/) directory. 

This document is written in [R markdown](http://rmarkdown.rstudio.com), which is a method of formatting text, code, and output to create documents that are sharable with other people. 
While this document is intended to serve as a reference for you to read while typing code into your own script,
you may also be interested in modifying and running code in the original R markdown file ([`week4.Rmd`](week4.Rmd) in the GitHub repository).
