---
title: 'Introduction to R, Class 2: Solutions'
output: github_document
---
  
<!--class2_solutions.md is generated from class2_solutions.Rmd. Please edit that file -->
  
  ```{r setup, include=FALSE, purl=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r include=FALSE}
library(tidyverse)
clinical <- read.csv("../data/clinical.csv")
```

#### Challenge-data
```{r eval=FALSE}
# example1: https://raw.githubusercontent.com/fredhutchio/R_intro/master/extra/clinical.tsv
download.file("https://raw.githubusercontent.com/fredhutchio/R_intro/master/extra/clinical.tsv", "data/clinical.tsv")
clinical_tsv <- read.delim("data/clinical.tsv")
# example2: https://raw.githubusercontent.com/fredhutchio/R_intro/master/extra/clinical.txt
download.file("https://raw.githubusercontent.com/fredhutchio/R_intro/master/extra/clinical.txt", "data/clinical.txt")
clinical_txt <- read.table("data/clinical.txt", header=TRUE)
```

It's also possible to read data into R directly using `read.csv(URL)`,
but it's often better practice to download data file for reference.

#### Challenge-extract
```{r}
# extract first column, assigning each result to object to show that data structure differs between output
test <- clinical[ , 1]
test2 <- clinical[1]
class(test) # factor (technically, a vector of factors)
class(test2) # data.frame
```

#### Challenge-days
```{r eval=FALSE}
clinical$days_to_death
clinical[6]
clinical[,6]
clinical[[6]]
clinical["days_to_death"] 
clinical[ , "days_to_death"] 
clinical[["days_to_death"]]
```

#### Challenge-rows
```{r}
clinical[1:6, c(3, 6)]
```

#### Challenge-calculate
```{r}
range(clinical$cigarettes_per_day, na.rm = TRUE)
mean(clinical$cigarettes_per_day, na.rm = TRUE)
```

#### Challenge-not-reported
```{r}
clinical <- read.csv("../data/clinical.csv", stringsAsFactors = TRUE)
ethnicity <- clinical$ethnicity
levels(ethnicity) # show levels of factor
# rename factors
levels(ethnicity)[3] <- NA
# replace race in data frame
clinical$ethnicity <- ethnicity
# check to see it worked
levels(ethnicity)
```

#### Challenge-remove
Perform a Google search to find additional strategies for renaming missing data.
For the search "replace missing data R",
the following result provides a lot of options: 
https://stackoverflow.com/questions/8161836/how-do-i-replace-na-values-with-zeros-in-an-r-dataframe

## Extra exercises

#### Challenge-disease-race
```{r}
# show index positions of column names
names(clinical)
# extract desired object
disease_race <- clinical[6733:6832, c(20, 16)]
```

#### Challenge-min-max
```{r}
min(clinical$days_to_death, na.rm = TRUE)
max(clinical$days_to_death, na.rm = TRUE)
range(clinical$days_to_death, na.rm = TRUE) # does both together
summary(clinical$days_to_death) # all summary stats for this answer
```

#### Challenge-factors
```{r}
# save column to object
race <- clinical$race 
# recall levels of object
levels(race) 
# rename factors
levels(race)[1] <- "American Indian"
levels(race)[2] <- "Asian"
levels(race)[3] <- "black"
levels(race)[4] <- "Pacific Islander"
levels(race)[5] <- NA
# recall levels of object
levels(race) 
# replace race in data frame
clinical$race <- race
```
